Using language as an indicator of cognitive decline in pre-clinical Alzheimer's Disease.

### 2019-12-05 Source/Paper 1 - Presidents Corpus Paper/PresidentsCorpus.tex
updated Pres

\section{Discussion}\label{discussion}
President Reagan received his diagnosis of AD in August 1994 but using transcripts of speeches he made in his two terms as President (January 1981 - January 1989) we have be able to identify certain changes in his use of language that we might ascribe to the onset of MCI and early AD. Despite differences in our methodology, our research supports some of the findings of Berisha and Liss in that we find an increase in non-specifc noun usage. Compared to our controls (GWHB and DJT), we find some slight trends with GWHB but no such trends with DJT in his speech albeit his samples of speech span a shorter amount of time. Interestingly, when we normalised the various types of words used by the presidents we found some interesting patterns that further differentiated RR from the controls. Whilst Non-specific nouns increased over time, we found that noun usage in general significantly decreased and pronouns increased similarly significantly. The increase in pronoun for those with early AD has been identified in literature, although there are only a few studies that explore this \cite{Almor1999, Wendelstein2015}. Wendlestein et al propose that the increased used of pronouns is an expression of an impaired ability to adapt language to the listener's needs \cite{Wendelstein2015}. Almor et al attributed this reliance on pronouns due to a impaired working memory \cite{Almor1999}.
\par 
The decrease in overall noun usage has also been identified as a feature. Jarrold et al found that AD patients would use more pronouns, verbs and fewer nouns than controls \cite{Jarrold2014}. Wendlestein in their investigations into noun usage found that decreased later on in AD progression and was unaffected in the pre-clinical stages of AD \cite{Wendelstein2014}. Our results are supported by existing literature and this potentially means that language analysis in the way we have structured it may have diagnostic or prognostic properties.
\par 
A criticism of Berisha and Liss's work is the problems they had with normalising the transcripts in terms of length. This was also a problem in the work of Garrard et al \cite{Garrard2005, Le2011}. Whilst it is important to control for outliers, there are other ways in which we can control for length of sample. In this paper, we controlled for transcript length by dividing any features that were raw counts by the total length of the transcript. When we did this, we found that there was a significant decrease in the number of unique words (R=-0.56, p\textless0.001) used however when we controlled via normalisation, we found that this was not a significant feature (R=-0.172, p=0.25). 
\par 
Overall, we are able to show differences in the a number of language features between RR, GHWB and DJT over time and the psychological literature confirms our findings in reference to these changes.
\par
Our next hypothesis involved exploring whether linear models were the most appropriate way to track this longitudinal data. In our analysis, we found there was no significant difference in terms of mean sum of squares errors between the linear model and the generalised additive model however this does not mean that these are equivalent. For example, considering the plot below. We compare a model fit to a linear model with a generalised additive model which aims to model the data more closely. It is clear to see that both models can track the movements of language use over time however, the linear model is quite a rigid model which does not allow for different inflection points. We can see that RR's decline is not linear, and that there are periods of relative stability and some periods where the decline is more severe. We argue that these more nuanced changes over time can only be modelled by a model such as a GAM.
\par
\begin{figure}[H]
	\centering
	\includegraphics[width=6cm, height=5cm]{plots/comparison1.png}
	\caption{Comparing a linear model with a generalised additive model}
\end{figure}

>>>>

\section{Discussion}\label{discussion}
President Reagan received his diagnosis of AD in August 1994 but using transcripts of speeches he made in his two terms as President (January 1981 - January 1989) we have be able to identify certain changes in his use of language that we might ascribe to the onset of MCI and early AD. Despite differences in our methodology, our research supports some of the findings of Berisha and Liss in that we find an increase in non-specifc noun usage. Compared to our controls (GWHB and DJT), we find some slight trends with GWHB but no such trends with DJT in his speech albeit his samples of speech span a shorter amount of time. Interestingly, when we normalised the various types of words used by the presidents we found some interesting patterns that further differentiated RR from the controls. Whilst Non-specific nouns increased over time, we found that noun usage in general significantly decreased and pronouns increased similarly significantly. The increase in pronoun for those with early AD has been identified in literature, although there are only a few studies that explore this \cite{Wendelstein2015}. Wendlestein et al propose that the increased used of pronouns is an expression of an impaired ability to adapt language to the listener's needs \cite{Wendelstein2015}. Almor et al attributed this reliance on pronouns due to a impaired working memory \cite{Almor1999}.
\par 
The decrease in overall noun usage has also been identified as a feature. Jarrold et al found that AD patients would use more pronouns, verbs and fewer nouns than controls \cite{Jarrold2014}. Wendlestein in their investigations into noun usage found that decreased later on in AD progression and was unaffected in the pre-clinical stages of AD \cite{Wendelstein2014}. Our results are supported by existing literature and this potentially means that language analysis in the way we have structured it may have diagnostic or prognostic properties.
\par 
A criticism of Berisha and Liss's work is the problems they had with normalising the transcripts in terms of length. This was also a problem in the work of Garrard et al \cite{Garrard2005, Le2011}. Whilst it is important to control for outliers, there are other ways in which we can control for length of sample. In this paper, we controlled for transcript length by dividing any features that were raw counts by the total length of the transcript. When we did this, we found that there was a significant decrease in the number of unique words (R=-0.56, p\textless0.001) used however when we controlled via normalisation, we found that this was not a significant feature (R=-0.172, p=0.25). 
\par 
Overall, we are able to show differences in the a number of language features between RR, GHWB and DJT over time and the psychological literature confirms our findings in reference to these changes.
\par
Our next hypothesis involved exploring whether linear models were the most appropriate way to track this longitudinal data. In our analysis, we found there was no significant difference in terms of mean sum of squares errors between the linear model and the generalised additive model however this does not mean that these are equivalent. For example, considering the plot below. We compare a model fit to a linear model with a generalised additive model which aims to model the data more closely. It is clear to see that both models can track the movements of language use over time however, the linear model is quite a rigid model which does not allow for different inflection points. We can see that RR's decline is not linear, and that there are periods of relative stability and some periods where the decline is more severe. We argue that these more nuanced changes over time can only be modelled by a model such as a GAM.
\par
\begin{figure}[H]
	\centering
	\includegraphics[width=6cm, height=5cm]{plots/comparison1.png}
	\caption{Comparing a linear model with a generalised additive model}
\end{figure}
